Supplementary Information

### 2021-08-09 doc/manuscript/ForestGEO_treerings.Rmd
changes in response to Chris Guiterman's review
#115

Understanding the climate sensitivity of tree growth is critical to predicting forest dynamics and productivity as the climate changes. 
The classic dendrochronological approach to characterizing the climate sensitivity of tree growth describes linear relationships between the primary growth-limiting climate factor (moisture or temperature) and population-level growth responses captured in ring-width index chronologies [@fritts_tree_1976; @speer_fundamentals_2010].
While invaluable for applications such as reconstructing past climates [e.g., @buntgen_2500_2011], accurately representing the climate sensitivity of forest productivity requires an analysis of sensitivity to multiple climatic variables and how this varies across trees in a forest stand [@babst_when_2018].
Precipitation and temperature can have additive or interactive effects on growth [@foster_predicting_2016; @meko_seascorr_2011;  @sanchez-salguero_disentangling_2015; @vlam_temperature_2014; @zuidema_recent_2020], and we hypothesize that both influence the growth of most species, often over different seasonal windows (Table `r hypotheses`).
In addition, we hypothesize that nonlinear climate responses, already known to be widespread within forest settings [@woodhouse_artificial_1999; @wilmking_global_2020; @rollinson_climate_2021], are in fact the predominant form of response in forests around the world (Table `r hypotheses`). 
Finally, the influence of DBH is typically removed through detrending [@cook_calculating_1997], which eliminates the potential to directly model its influence on climate sensitivity, but we hypothesize that interactive effects of DBH and climate are, in fact, quite common in forest settings and fundamental to understanding climate change responses of forests [Table 1, @bennett_larger_2015; @trouillier_size_2019; @mcgregor_tree_2021; @rollinson_climate_2021].

Tree DBH scales with numerous traits affecting tree growth (e.g., height, crown size and position, root mass, hydraulic architecture) and therefore is itself linked to growth [e.g., @anderson-teixeira_sizerelated_2015].
Yet, there remain inconsistencies across disciplines (dendrochronology, forest ecology) as to what is considered a typical growth pattern. 
Dendrochronological records from trees that established in high-light environments commonly show a pattern in which radial stem growth increments (ring width, RW) are initially large and decline with DBH [@fritts_tree_1976], whereas censuses of globally distributed forests indicate that diameter increments most commonly increase with DBH [@muller-landau_testing_2006; @anderson-teixeira_sizerelated_2015].
We hypothesize that this discrepancy is primarily a distinction between trees that establish in the open versus in the understory [Table `r hypotheses`, @lorimer_estimating_1988].
Building upon observed ontogenetic patterns in RW, dendrochronology studies often adopt a null hypothesis that the annual basal area increment (BAI) fluctuates around a constant mean after a juvenile growth phase [@fritts_tree_1976; @biondi_theorydriven_2008] -- a pattern that we would not expect to hold in understory-established trees (Table `r hypotheses`).
Finally, there is debate as to whether the aboveground biomass increment ($\Delta$AGB) increases continuously with DBH [@stephenson_rate_2014; @meakem_role_2018; @foster_predicting_2016; @sillett_increasing_2010b] or peaks at intermediate DBH and then plateaus or declines as trees divert carbon to other functions such as reproduction and respiration [@thomas_agerelated_2011; @sheil_does_2017; @west_increasing_2020]. 
Following the finding that the latter pattern is common for individual trees whereas the former emerges in "cross-sectional" analyses of forest stands [@forrester_does_2021], we hypothesize that $\Delta$AGB -- and also BAI -- peaks and declines as DBH increases (Table `r hypotheses`).
Discerning these ontogenetic growth trends is essential not only for predicting the growth rate of any given tree, but also for standardizing for DBH to deduce the influence of slowly changing environmental drivers [see next paragraph, @peters_detecting_2015], with the reliability of such analyses contingent upon accurate assumptions of ontogenetic growth patterns.

@@

All tree cores were cross-dated and measured by the original researchers using standard dendrochronological practices [@stokes_introduction_1968]. 
From among the full set of original RW measurements, we excluded cores for which we detected technical errors (e.g., labeling inconsistencies, obvious dating errors) that could not be resolved before finalizing the analysis. 
We also excluded records with small sample sizes or highly anomalous growth patterns, including
(1) species with < 7 cores,
(2) cores with < 30 years of record,
(3) contiguous portions of cores containing large outliers (RW > mean plus 5 x SD of RW for the entire core),
and (4) the final 20 years prior to death for trees cored dead. 
The final criterion was implemented to avoid periods of growth decline and potentially altered climate sensitivity prior to death [@cailleret_synthesis_2017; @desoto_low_2020].
From analyses including DBH (see below), we further excluded
(1) trees for which we lacked data required to reconstruct DBH,
(2) trees for which there was a significant inconsistency between measured DBH and the sum of RW's across the core (Appendix `r DBH_reconstruction`),
and (3) poorly represented tails of the DBH distribution, starting where reconstructed DBH (see below) included < 3 conspecific trees.
In total, this resulted in inclusion of `r n_cores_all` cores from `r n_trees_all` trees, `r n_cores_dbh` of which (from `r n_trees_dbh` trees) could be included in analyses with DBH (Table `r sampling_details_table`).

@@

Across diverse climates and forest types (Table `r sites`), growth rates of `r n_species` tree species usually responded positively to water availability (PPT or PDF) -- at least up until the long-term mean -- 
and negatively to temperature (usually $T_{max}$ or PET), with the exception of several positive responses at times and in places where temperature was limiting (Table `r hypotheses`, Fig. `r clim_sens`).
These findings are generally consistent with current understanding of global-scale patterns in climate sensitivity [@rozendaal_dendroecology_2011; @babst_twentieth_2019]: outside of the wet tropics (where there are few tree-ring records), the majority of forests are moisture limited and respond negatively to temperature, with a shrinking area of temperature-limited forests in cold, humid regions (with SC falling near the transition zone). 
Within warmer regions, warm winter or early spring temperatures in humid climates may advance the growing season [@keenan_net_2014] and increase annual growth [@babst_twentieth_2019; @tumajer_increasing_2017], as we show for all three species at ZOF and one species at HF (Fig. `r clim_sens`).
However, the predominantly negative temperature responses (Fig. `r clim_sens`) imply that warmer temperatures are likely to reduce growth across the wide range of forest types and climates represented here.
The primary mechanism underlying growth decreases at high temperatures is presumably increased evaporative demand (PET or VPD) and ensuant exacerbation of observed water limitations [@novick_increasing_2016; @humphrey_soil_2021; @lopez_systemic_2021].
This effect occurs in addition to the effects of precipitation (Fig. `r clim_sens`), highlighting the fact that temperature and associated VPD increases limit growth even under conditions of high soil moisture [@novick_increasing_2016], and occurs over shorter time-frames (usually $\le$ 3 mo) than the effects of precipitation (usually $\ge$ 3 mo.; Table `r hypotheses`, Fig. `r clim_sens`).
This suggests that relatively short periods of anomalously high temperatures and evaporative demand, themselves caused in large part by soil dryness [@humphrey_soil_2021], add to effects of prolonged periods of reduced precipitation to shape forest drought responses.

@@

Our finding that BAI and $\Delta$AGB generally saturate or decline with increasing DBH (Table `r hypotheses`, Fig. `r size`) contrasts with findings of cross-sectional analyses of forest census data showing that $\Delta$AGB increases continuously with DBH [@stephenson_rate_2014; @meakem_role_2018].
In large part, this discrepancy can be explained by differences between cross-sectional analyses and "longitudinal" patterns of individual trees through time [@sheil_does_2017; @forrester_does_2021].
Declines in BAI and $\Delta$AGB at larger DBH are probably in part attributable to increasing allocation to reproduction [@thomas_agerelated_2011], and are also linked to slowly changing environmental conditions (e.g., successional changes in stand structure, climate change).
Apparent declines in $\Delta$AGB at large DBH (or old age) may also be driven by shifts towards proportionally greater wood production within the crown [e.g., branch production, @sillett_increasing_2010b; @sillett_comparative_2021] that are not adequately captured by biomass allometries based on DBH and sometimes height [@goodman_importance_2014b; @disney_new_2020].
Notably, inclusion of year in the GLS models tended to reduce the magnitude of BAI and $\Delta$AGB declines at larger DBH (Figs. `r GLS_SCBI`, `r GLS_SCBI_year`), suggesting that some of the declines at large DBH (Fig. `r size`) are more properly attributed to recent environmental changes than to large DBH.

>>>>

Understanding the climate sensitivity of tree growth is critical to predicting forest dynamics and productivity as the climate changes. 
The classic dendrochronological approach to characterizing the climate sensitivity of tree growth describes linear relationships between the primary growth-limiting climate factor (moisture or temperature) and population-level growth responses captured in ring-width index chronologies [@fritts_tree_1976; @cook_time_1985; @speer_fundamentals_2010].
While invaluable for applications such as reconstructing past climates [e.g., @buntgen_2500_2011], accurately representing the climate sensitivity of forest productivity requires an analysis of sensitivity to multiple climatic variables and how this varies across trees in a forest stand [@babst_when_2018].
Precipitation and temperature can have additive or interactive effects on growth [@foster_predicting_2016; @meko_seascorr_2011;  @sanchez-salguero_disentangling_2015; @vlam_temperature_2014; @zuidema_recent_2020], and we hypothesize that both influence the growth of most species, often over different seasonal windows (Table `r hypotheses`).
In addition, we hypothesize that nonlinear climate responses, already known to be widespread within forest settings [@woodhouse_artificial_1999; @wilmking_global_2020; @rollinson_climate_2021], are in fact the predominant form of response in forests around the world (Table `r hypotheses`). 
Finally, the influence of DBH is typically removed through detrending [@cook_calculating_1997], which eliminates the potential to directly model its influence on climate sensitivity, but we hypothesize that interactive effects of DBH and climate are, in fact, quite common in forest settings and fundamental to understanding climate change responses of forests [Table 1, @bennett_larger_2015; @trouillier_size_2019; @mcgregor_tree_2021; @rollinson_climate_2021].

Tree DBH scales with numerous traits affecting tree growth (e.g., height, crown size and position, root mass, hydraulic architecture) and therefore is itself linked to growth [e.g., @anderson-teixeira_sizerelated_2015].
Yet, there remain inconsistencies across disciplines (dendrochronology, forest ecology) as to what is considered a typical growth pattern. 
Dendrochronological records from trees that established in high-light environments commonly show a pattern in which radial stem growth increments (ring width, RW) are initially large and decline with DBH [@fritts_tree_1976], whereas censuses of globally distributed forests indicate that diameter increments most commonly increase with DBH [@muller-landau_testing_2006; @anderson-teixeira_sizerelated_2015].
We hypothesize that this discrepancy is primarily a distinction between trees that establish in the open versus in the understory [Table `r hypotheses`, @lorimer_estimating_1988].
Building upon observed ontogenetic patterns in RW, dendrochronology studies often adopt a null hypothesis that the annual basal area increment (BAI) fluctuates around a constant mean after a juvenile growth phase [@fritts_tree_1976; @biondi_theorydriven_2008] -- a pattern that we would not expect to hold in understory-established trees (Table `r hypotheses`).
Finally, there is debate as to whether the aboveground biomass increment ($\Delta$AGB) increases continuously with DBH [@stephenson_rate_2014; @meakem_role_2018; @foster_predicting_2016; @sillett_increasing_2010] or peaks at intermediate DBH and then plateaus or declines as trees divert carbon to other functions such as reproduction and respiration [@thomas_agerelated_2011; @sheil_does_2017; @west_increasing_2020]. 
Following the finding that the latter pattern is common for individual trees whereas the former emerges in "cross-sectional" analyses of forest stands [@forrester_does_2021], we hypothesize that $\Delta$AGB -- and also BAI -- peaks and declines as DBH increases (Table `r hypotheses`).
Discerning these ontogenetic growth trends is essential not only for predicting the growth rate of any given tree, but also for standardizing for DBH to deduce the influence of slowly changing environmental drivers [see next paragraph, @peters_detecting_2015], with the reliability of such analyses contingent upon accurate assumptions of ontogenetic growth patterns.

@@

All tree cores were cross-dated and measured by the original researchers using standard dendrochronological practices [@stokes_introduction_1968]. 
From among the full set of original RW measurements, we excluded cores for which we detected technical errors (e.g., labeling inconsistencies, obvious dating errors) that could not be resolved before finalizing the analysis. 
We also excluded records with small sample sizes or highly anomalous growth patterns, including
(1) species with < 7 cores,
(2) cores with < 30 years of record,
(3) contiguous portions of cores containing large outliers (RW > mean plus 5 x SD of RW for the entire core),
and (4) the final 20 years prior to death for trees cored when dead. 
The final criterion was implemented to avoid periods of growth decline and potentially altered climate sensitivity prior to death [@cailleret_synthesis_2017; @desoto_low_2020].
From analyses including DBH (see below), we further excluded
(1) trees for which we lacked data required to reconstruct DBH,
(2) trees for which there was a significant inconsistency between measured DBH and the sum of RW's across the core (Appendix `r DBH_reconstruction`),
and (3) poorly represented tails of the DBH distribution, starting where reconstructed DBH (see below) included < 3 conspecific trees.
In total, this resulted in inclusion of `r n_cores_all` cores from `r n_trees_all` trees, `r n_cores_dbh` of which (from `r n_trees_dbh` trees) could be included in analyses with DBH (Table `r sampling_details_table`).

@@

Across diverse climates and forest types (Table `r sites`), growth rates of `r n_species` tree species usually responded positively to water availability (PPT or PDF) -- at least up until the long-term mean -- 
and negatively to temperature (usually $T_{max}$ or PET), with the exception of several positive responses at times and in places where temperature was limiting (Table `r hypotheses`, Fig. `r clim_sens`).
These findings are generally consistent with current understanding of global-scale patterns in climate sensitivity [@rozendaal_dendroecology_2011; @babst_twentieth_2019]: outside of the wet tropics (where there are few tree-ring records), the majority of forests are moisture limited and respond negatively to temperature, with a shrinking area of temperature-limited forests in cold, humid regions (with SC falling near the transition zone). 
Within warmer regions, warm winter or early spring temperatures in humid climates may advance the growing season [@keenan_net_2014] and increase annual growth [@babst_twentieth_2019; @tumajer_increasing_2017], as we show for all three species at ZOF and one species at HF (Fig. `r clim_sens`).
However, the predominantly negative temperature responses (Fig. `r clim_sens`) imply that warmer temperatures are likely to reduce growth across the wide range of forest types and climates represented here.
The primary mechanism underlying growth decreases at high temperatures is presumably increased evaporative demand (PET or vapor pressure deficit, VPD) and ensuant exacerbation of observed water limitations [@novick_increasing_2016; @humphrey_soil_2021; @lopez_systemic_2021].
This effect occurs in addition to the effects of precipitation (Fig. `r clim_sens`), highlighting the fact that temperature and associated VPD increases limit growth even under conditions of high soil moisture [@novick_increasing_2016], and occurs over shorter time-frames (usually $\le$ 3 mo) than the effects of precipitation (usually $\ge$ 3 mo.; Table `r hypotheses`, Fig. `r clim_sens`).
This suggests that relatively short periods of anomalously high temperatures and evaporative demand, themselves caused in large part by soil dryness [@humphrey_soil_2021], add to effects of prolonged periods of reduced precipitation to shape forest drought responses.

@@

Our finding that BAI and $\Delta$AGB generally saturate or decline with increasing DBH (Table `r hypotheses`, Fig. `r size`) contrasts with findings of cross-sectional analyses of forest census data showing that $\Delta$AGB increases continuously with DBH [@stephenson_rate_2014; @meakem_role_2018].
In large part, this discrepancy can be explained by differences between cross-sectional analyses and "longitudinal" patterns of individual trees through time [@sheil_does_2017; @forrester_does_2021].
Declines in BAI and $\Delta$AGB at larger DBH are probably in part attributable to increasing allocation to reproduction [@thomas_agerelated_2011], and are also linked to slowly changing environmental conditions (e.g., successional changes in stand structure, climate change).
Apparent declines in $\Delta$AGB at large DBH (or old age) may also be driven by shifts towards proportionally greater wood production within the crown [e.g., branch production,  @sillett_comparative_2021; @sillett_increasing_2010] that are not adequately captured by biomass allometries based on DBH and sometimes height [@goodman_importance_2014b; @disney_new_2020].
Notably, inclusion of year in the GLS models tended to reduce the magnitude of BAI and $\Delta$AGB declines at larger DBH (Figs. `r GLS_SCBI`, `r GLS_SCBI_year`), suggesting that some of the declines at large DBH (Fig. `r size`) are more properly attributed to recent environmental changes than to large DBH.
